Well-quasi-orderings on word languages

### 2025-09-08 src/amalgam.tex
new: add refs + morhpic words + todos

\AP In this section, we are going to design an effective decision procedure for
\kl{well-quasi-ordering} by the \kl{infix relation}. To that end, the first
requirement is to fix a way to represent languages $L \subseteq \Sigma^*$.
Traditionally, one would use finite automata, context-free grammars, or for the
more adventurous, vector addition systems with states. However, our proof
technique will only require us to have a way to ``glue'' together runs of the
system to ``pump'' them and produce new runs: this is the usual pumping lemma
in automata theory, and Ogden's lemma for context-free grammars \cite{OGDEN68}.
It turns out that there is a rather large family of systems for which pumping
arguments based on so-called \emph{minimal runs} exist: they are called
\emph{amalgamation systems} and were recently introduced by Anand, Schmitz,
Schütze, and Zetzsche \cite{ASZZ24}.

>>>>

\AP In the previous section, we have represented languages that are
\kl{downwards closed} by the \kl{infix relation} as \kl{infixes} of infinite
words. However, there are many other natural ways to represent languages, such
as finite automata or context-free grammars. In this section, we are going to
show that our results on \kl{bounded languages} can be applied to a large class
of systems, called \kl{amalgamation systems}, that includes as particular
examples finite automata and context-free grammars.

### 2025-09-08 src/infixes.tex
new: add refs + morhpic words + todos

\begin{corollary}
    \label{ordinal-invariants-bounded:cor}
    Let $L$ be a \kl{bounded language} of $\Sigma^*$
    that is \kl{well-quasi-ordered} by the \kl{infix relation}.
    Then, the \kl{ordinal width} of $L$ less than $\omega^2$,
    its \kl{ordinal height} is at most $\omega$,
    and its \kl{maximal order type} less than $\omega^3$.
    Furthermore, those three bounds are tight.
\end{corollary}
\begin{proof}
  Upper bounds are a direct consequence of \cref{bounded-language:thm},
  and the tightness is witnessed by the 
  languages: 
  $L_k \defined \bigcup_{i = 2}^{k+1} (a b^i a)^* (b a^i b)^*$,
  that are \kl{bounded languages} of $\set{a,b}^*$,
  \kl{well-quasi-ordered} by the \kl{infix relation},
  and have \kl{ordinal width}, \kl{ordinal height} and
  \kl{maximal order type} respectively equal to $\omega \cdot k$, $\omega$ and $\omega^2 \cdot k$.
\end{proof}

@@

\subsection{Downwards Closures and Bounded Languages}

Let us now discuss the implications of this characterization in terms of
\kl{downwards closures}: if $L$ is a \kl{bounded language}, then considering
$L$ or its \kl{downwards closure} is equivalent with respect to being
\kl{well-quasi-ordered} by the \kl{infix relation}.

@@

\begin{theorem}
  \label{bounded-wqo-dw-closed-regular:thm}
  Let $L$ be a \kl{downwards closed} and \kl{well-quasi-ordered} language
  of $\Sigma^*$ for the \kl{infix relation}. The following are equivalent:
  \begin{enumerate}
    \item $L$ is a \kl{bounded language},
    \item $L$ is a \kl{regular language},
    \item There exists a finite set $E \subset_{\mathsf{fin}} (\Sigma^*)^3$ such that
      $L = \bigcup_{(x,u,y) \in E} \InfPeriodChain{x} u \InfPeriodChain{y}$.
  \end{enumerate}
\end{theorem}

\subsection{General Downwards Closed Languages}

@@

\todo[inline]{aliaume: write it}
Let us conclude this section with a few remarks on the notion of
\kl{factor complexity} of languages.

\begin{lemma}
  There exists a language $L$ with exponential factor complexity 
  that is \kl{well-quasi-ordered} for the \kl{infix relation}.
\end{lemma}

@@

Note that being \kl{uniformly recurrent} still leaves open a lot of potential
behaviours, for instance, Sturmian words are \kl{uniformly recurrent}, but are
not \kl(sequence){automatic}. Interestingly enough, the \kl{ordinal invariants}
of the \kl{automatic sequences} are not smaller than the ones of \kl{ultimately
uniformly recurrent} words: this contrasts with the case of languages that are
\kl{well-quasi-ordered} for the \kl{infix relation} in general, where the
hypothesis of being representable (by a \kl{regular language}, or an
\kl{amalgamation system}) will drastically reduce the \kl{ordinal invariants}
of said language (\cref{ordinal-invariants-bounded:cor}, \cref{infix-amalgamation:thm}).

\todo[inline]{For Nathan: 
  talk about morphic words / sequences and 
  decidability here.
  \url{https://arxiv.org/pdf/1204.5393}
}

\todo[inline]{Talk about factor complexity and 
well-quasi-orderings: the two are unrelated. This can be a remark.
\url{https://arxiv.org/pdf/2206.00376}
Here example 5 p8 is an explicit example of a uniformly recurrent word with exponential factor complexity.
}

>>>>

\begin{corollary}
    \label{ordinal-invariants-bounded:cor}
    Let $L$ be a \kl{bounded language} of $\Sigma^*$
    that is \kl{well-quasi-ordered} by the \kl{infix relation}.
    Then, the \kl{ordinal width} of $L$ less than $\omega^2$,
    its \kl{ordinal height} is at most $\omega$,
    and its \kl{maximal order type} less than $\omega^3$.
    Furthermore, those three bounds are tight.
\end{corollary}
\begin{proof}
  Upper bounds are a direct consequence of \cref{bounded-language:thm},
  and the tightness is witnessed by the 
  languages: 
  $L_k \defined \bigcup_{i = 2}^{k+1} (a b^i a)^* (b a^i b)^*$,
  that are \kl{bounded languages} of $\set{a,b}^*$,
  \kl{well-quasi-ordered} by the \kl{infix relation},
  and have \kl{ordinal width}, \kl{ordinal height} and
  \kl{maximal order type} respectively equal to $k \cdot \omega$, $\omega$ and $k \cdot \omega^2$.
\end{proof}

@@

Let us now discuss another classical restriction that can be imposed on 
languages when studying \kl{well-quasi-orders}, that of being
\kl{downwards closed}. Indeed, the \cref{infix-embedding:thm}
crucially relies on constructing languages that are \emph{not}
\kl{downwards closed}, and we have shown 
in \cref{dw-closure-not-wqo:rem} that the \kl{downwards closure}
of a \kl{well-quasi-ordered} language is not necessarily
\kl{well-quasi-ordered}.

\subsection{Characterization of Well-Quasi-Ordered Downwards Closed Languages}

An immediate consequence of \cref{bounded-language:thm} is that   if $L$ is a
\kl{bounded language}, then considering $L$ or its \kl{downwards closure}
$\dwset[\infleq]{L}$ is equivalent with respect to being
\kl{well-quasi-ordered} by the \kl{infix relation}, as opposed to
the general case illustrated in \cref{dw-closure-not-wqo:rem}.

@@

@@

\AP
Let us conclude this section with a few remarks on the notion of \kl{factor
complexity} of languages. Recall that the \intro{factor complexity} of a
language $L \subseteq \Sigma^*$ is the function $f_L: \Nat \to \Nat$ such that
$f_L(n)$ is the number of distinct words of size $n$ in $L$. A language $L$ has
\intro{exponential factor complexity} if there exists a constant $C > 1$ such
that $f_L(n) \geq C^n$ for infinitely many $n \in \Nat$.

\begin{lemma}
  \label{exponential-factor-complexity:lem}
  There exists a language $L$ with \kl{exponential factor complexity}
  that is \kl{well-quasi-ordered} for the \kl{infix relation}.
\end{lemma}
\begin{proof}
  We leverage results from \cite{CAKA97} on Toeplitz words.
  Namely, the 
  $(5, 3)$-Toeplitz word is \kl{uniformly recurrent} 
  \cite[p. 499]{CAKA97},
  but has an exponential factor complexity 
  \cite[Theorem 5]{CAKA97}.
\end{proof}

It is the case that \kl{bounded languages} have at most a polynomial factor
complexity, and that the \kl{Thue-Morse sequence} has a linear factor
complexity. More generally, as we will see, most ``concrete'' examples of
infinite words have a low factor complexity: \kl{automatic sequences} have at
most a linear factor complexity, and \kl(morphic){morphic sequences} have at
most a polynomial factor complexity. In that sense, the
\cref{exponential-factor-complexity:lem} shows that one can still construct
``pathological'' examples of infinite words that are \kl{well-quasi-ordered}
for the \kl{infix relation}, but have a high factor complexity.

@@

One can go beyond \cref{automatic-wqo:thm} and consider
\kl(morphic){morphic sequences}, which are obtained by iterating a
\kl(substitution){substitution} on a finite alphabet, and then applying a
\kl(homomorphism){homomorphism}. The \kl{Fibonacci sequence} is an example of
a \kl(morphic){morphic sequence}. It is known that one can decide whether a
\kl(morphic){morphic sequence} is \kl{uniformly recurrent}
\cite[Theorem 1]{DURAND13}.

\todo[inline]{Define properly morphic words and 
  the questions associated to them. Also, there is a difference between
  \kl{ultimately uniformly recurrent} and \kl{uniformly recurrent}
  that is not handled here.}
\begin{theorem}
    \label{morphic-wqo:thm}
    Given a \kl(morphic){morphic sequence} $w \in \Sigma^{\Nat}$, one can decide
    whether the set of \kl{infixes} of $w$ is \kl{well-quasi-ordered} for the
    \kl{infix relation}.
\end{theorem}
